---
title: "Reproducible Research: Peer Assessment 1"
author: "R.M. Palmer"
date: "March 18, 2017"
output: 
  html_document:
    keep_md: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = 'figure/')
```


## Loading and preprocessing the data
```{r load_data}
library(plyr)
library(ggplot2)
library(chron)

unzip(zipfile="activity.zip")
raw_data <- read.csv("activity.csv")

```

## What is mean total number of steps taken per day?
```{r compute_original}
steps_per_day <- with(raw_data,tapply(steps,date,sum,na.rm=TRUE))
hist(steps_per_day,breaks=25,col="light blue",main="Histogram of steps in original data")
mean_steps_per_day <- mean(steps_per_day)
median_steps_per_day <- median(steps_per_day)
```

The mean number of steps per day is `r sprintf("%6.1f",mean_steps_per_day)`
The median number of steps per day is `r sprintf("%6d",median_steps_per_day)`

## What is the average daily activity pattern?
```{r average_intervals}
interval_means <- with(raw_data,tapply(steps,interval,mean,na.rm=TRUE))
interval <- unique(raw_data$interval)
interval_plot <- qplot(interval,interval_means) + 
  xlab("interval") + 
  ylab("steps") + 
  geom_line() + 
  ggtitle("Steps per interval")
print(interval_plot)
peak_interval <- which.max(interval_means)
peak_time <- names(peak_interval)
peak_index <- peak_interval[[1]]
```

The interval with maximum number of steps, averaged across all days in the datasets, is `r peak_time`, the `r peak_index`-th interval

## Imputing missing values
```{r imput_missing,results="hide"}
missing_count = sum(is.na(raw_data$steps))
my_lookup <- data.frame(interval,interval_means)
infilled <- merge(raw_data,my_lookup)
infilled <- mutate(infilled,steps=ifelse(is.na(steps),interval_means,steps))
```

There are `r missing_count` missing values.

After replacing missing values with the mean over days by interval, a histogram looks like the following:

```{r imputed_stats}
new_steps_per_day <- with(infilled,tapply(steps,date,sum,na.rm=TRUE))
hist(new_steps_per_day,breaks=25,col="light blue",main="Histogram of steps in infilled data")
new_mean_steps_per_day <- mean(new_steps_per_day)
new_median_steps_per_day <- median(new_steps_per_day)

```

After replacing missing values,  

The new mean number of steps per day is `r sprintf("%6.1f",new_mean_steps_per_day)`
The new median number of steps per day is `r sprintf("%6.1f",new_median_steps_per_day)` 

Both the mean and median were raised after imputing missing data in this way.

## Are there differences in activity patterns between weekdays and weekends?
```{r weekend_or_not}

labeled <- mutate(raw_data,w =as.factor(ifelse(is.weekend(date),"weekend","weekday")))
breakdown <- with(labeled,tapply(steps,list("w"=w,"interval"=interval),mean,na.rm=TRUE))
par(mfrow=c(2,1))
plot(interval,breakdown[1,],type="l",xlab="interval",ylab="steps",main="Weekday Pattern")
plot(interval,breakdown[2,],type="l",xlab="interval",ylab="steps",main="Weekend Pattern")
```

On weekdays, there is a higher level of activity in the mornings.  On weekends, the overall activity level is higher and more varried.
